---
title: "Global Earthquake Activity"
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

clean <- read_csv("data/clean/earthquakes-clean.csv",
                   show_col_types = FALSE)

strongest <- clean |> slice_max(mag, n = 1)
```

## Summary

In the past 30 days, USGS recorded `r nrow(clean)` earthquakes of magnitude 4.5 or greater worldwide. The strongest was a magnitude `r strongest$mag` event near `r strongest$place` on `r format(strongest$date, "%B %d, %Y")`. The median magnitude was `r round(median(clean$mag), 1)`, and `r sum(clean$mag >= 6.0)` earthquakes reached magnitude 6.0 or higher.

## Strongest Earthquakes

```{r}
#| label: top-table
clean |>
  slice_max(mag, n = 10) |>
  mutate(
    date = format(date, "%b %d"),
    depth = paste0(round(depth), " km"),
    mag = round(mag, 1)
  ) |>
  select(Date = date, Magnitude = mag, Depth = depth, Location = place) |>
  knitr::kable()
```

## Earthquakes Over Time

```{r}
#| label: timeline-plot
#| fig-width: 9
#| fig-height: 4
daily <- clean |>
  count(date)

ggplot(daily, aes(x = date, y = n)) +
  geom_col(fill = "#2c3e50", alpha = 0.8) +
  labs(
    title = "Earthquakes Per Day (M4.5+)",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

## Magnitude Distribution

```{r}
#| label: mag-histogram
ggplot(clean, aes(x = mag)) +
  geom_histogram(binwidth = 0.2, fill = "#1f9ac9", color = "white") +
  labs(
    title = "Magnitude Distribution",
    x = "Magnitude",
    y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

## Where They Happened

```{r}
#| label: map-plot
#| fig-width: 10
#| fig-height: 5
world <- map_data("world")

ggplot() +
  geom_polygon(
    data = world,
    aes(x = long, y = lat, group = group),
    fill = "grey90", color = "grey80", linewidth = 0.2
  ) +
  geom_point(
    data = clean,
    aes(x = longitude, y = latitude, size = mag, color = mag),
    alpha = 0.6
  ) +
  scale_color_viridis_c(option = "inferno", end = 0.9) +
  scale_size_continuous(range = c(0.5, 4)) +
  labs(
    title = "Global Earthquake Locations (Past 30 Days, M4.5+)",
    color = "Magnitude",
    size = "Magnitude"
  ) +
  coord_fixed(1.3) +
  theme_void(base_size = 13) +
  theme(legend.position = "bottom")
```

## Most Active Regions

```{r}
#| label: region-table
clean |>
  count(region, sort = TRUE) |>
  slice_head(n = 10) |>
  rename(Region = region, Earthquakes = n) |>
  knitr::kable()
```


